---
title: "BIOS7345 Lab 7"
subtitle: "Factor coding schemes"
author: "Sarah Lotspeich"
date: "2 November 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, tidy = TRUE)

#Load packages
library(magrittr)
```

#Introduction
We are interested in comparing the `grades` of students in 3 different mathematics `classes`. A sample of 30 students is obtained.

```{r}
set.seed(12)
n <- 30 #number of students
grades <- rnorm(n, 75, 5)
classes <- sample(c("Class A", "Class B", "Class C"), n, replace = TRUE) %>% factor(levels = c("Class A", "Class B", "Class C"))
```

Begin by finding the overall mean grade across all classes 

```{r}
#overall mean grade
```

as well as the mean grade in each class

```{r}
#class-specific mean grades
aggregate()
```

Before we begin, a [helpful lil page](https://stats.idre.ucla.edu/r/library/r-library-contrast-coding-systems-for-categorical-variables/) on factor coding schemes.

#Model 1: Reference cell coding
Reference cell coding is what we commonly refer to as "dummy variable coding". Since we have $k = 3$ levels of the `classes` variable, how many indicator variables will we need? 

There are three levels to the unordered factor variable `classes`

```{r}
#inspect the cell counts in each class

```

so we will have 3 - 1 = 2 contrasts.

##Fit the model
```{r}
options(contrasts = c("contr.treatment", "contr.poly"))
mod <- lm(grades ~ classes)
summary(mod)
```

What does your $\pmb{\beta}$ vector look like for this model? 

$$\pmb{\beta} = \begin{bmatrix}\beta_0 \\ \beta_1 \\ \beta_2\end{bmatrix}$$

##Design matrix
Obtain the design matrix for this coding scheme (ignoring replicates) from your  `mod` object.  

```{r}
#save the unique rows of your design matrix

```

Stop and think: which group is your reference group? 

\begin{quote}
The reference group is identified as that for which all indicator variables are equal to 0. In our case, if we assume that the first row is Class A then Class A is our reference group. 
\end{quote}

##Group means
We can obtain the class-specific means from the model parameter estimates by multiplying the unique rows of the design matrix (`X`) by the coefficients. 

```{r}
#obtain class-specific means 
```

How do these means compare to those we calculated initially using `aggregate()`? 

\begin{quote}

\end{quote}

So what does this mean? When using reference cell coding, we have 

  1.  $\bar{X}_1 =$ 
  2.  $\bar{X}_2 =$ 
  3.  $\bar{X}_3 =$ 
  
If we knew the true mean grades in each class ($\mu_A, \mu_B, \mu_C$, respectively), how could we obtain the model parameters directly? 

  1.  $\beta_0 = $: 
  2.  $\beta_1 = $: 
  3.  $\beta_2 = $: 
  
#Model 2: deviation coding
Whereas reference cell coding compared each level to a reference level, deviation coding compares each level to a grand mean. 

##Fit the model
```{r}
options(contrasts = c("contr.sum", "contr.poly"))
mod <- lm(grades ~ classes)
summary(mod)
```

What does your $\pmb{\beta}$ vector look like for this model? 

$$\pmb{\beta} = \begin{bmatrix}\mu \\ \alpha_1 \\ \alpha_2\end{bmatrix}$$
Here, $\mu$ is the \underline{grand mean}... but what's the grand mean anyway? 

```{r}
#is it the overall mean? 


#or is it the mean of the means of the dependent variable at all levels of the factor

```

##Design matrix
Obtain the design matrix for this coding scheme (ignoring replicates) from your  `mod` object.

```{r}
#save the unique rows of your design matrix

```

How do we interpret the rows of `X`? 

  1.  
  2.  
  3.  

##Group means
We can obtain the class-specific means from the model parameter estimates by multiplying the unique rows of the design matrix (`X`) by the coefficients. 

```{r}
#obtain class-specific means

```

How do these means compare to those we calculated initially using `aggregate()`? 

\begin{quote}

\end{quote}

So what does this mean? When using deviation coding, we have 

  1.  $\bar{X}_1 =$
  2.  $\bar{X}_2 =$
  3.  $\bar{X}_3 =$
  
If we knew the true mean grades in each class ($\mu_A, \mu_B, \mu_C$, respectively) and the true grand mean ($\bar{\mu}$), how could we obtain the model parameters directly? 

  1.  $\mu = $: 
  2.  $\alpha_1 = $: 
  3.  $\alpha_2 = $: 
  
#Model 3: Helmert coding
Whereas reference cell coding compared each level to a reference level and deviation coding compares each level to a grand mean, Helmert coding compares each level with the mean of the subsequent levels. 

##Fit the model
```{r}
options(contrasts = c("contr.helmert", "contr.poly"))
mod <- lm(grades ~ classes)
summary(mod)
```

What does your $\pmb{\beta}$ vector look like for this model? 

$$\pmb{\beta} = \begin{bmatrix}\beta_0 \\ \beta_1 \\ \beta_2\end{bmatrix}$$

##Design matrix
Obtain the design matrix for this coding scheme (ignoring replicates) from your  `mod` object.

```{r}
#save the unique rows of your design matrix

```

How do we interpret the rows of `X`? 

  1.  
  2.  
  3.  

##Group means
We can obtain the class-specific means from the model parameter estimates by multiplying the unique rows of the design matrix (`X`) by the coefficients. 

```{r}
#obtain class-specific means

```

How do these means compare to those we calculated initially using `aggregate()`? 

\begin{quote}

\end{quote}

So what does this mean? When using deviation coding, we have 

  1.  $\bar{X}_1 = $
  2.  $\bar{X}_2 = $
  3.  $\bar{X}_3 = $
  
If we knew the true mean grades in each class ($\mu_A, \mu_B, \mu_C$, respectively) and the true grand mean ($\bar{\mu}$), how could we obtain the model parameters directly? Interpret.

  1.  $\beta_0 = $: 
  2.  $\beta_1 = $:
  3.  $\beta_2 = $: